---
title: "Homework Assignment 1"
author: "Arshaq Razack"
date: 'Assigned: Oct 24, 2020, Due Sun Nov 01, 2020 11:59PM'
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
---

# Data Exploration

```{r}
nyc = read.csv("nycflights.csv")

# Delete first column
nyc$X = NULL

# Factorizing variables 
nyc$origin = as.factor(nyc$origin)
nyc$tailnum = as.factor(nyc$tailnum)
nyc$month = as.factor(nyc$month)
nyc$dest = as.factor(nyc$dest)
nyc$carrier = as.factor(nyc$carrier)
nyc$flight = as.factor(nyc$flight)

```


- How many airlines are there?

There are `r length(levels(nyc$carrier))` airlines

- How many flights there were by the airline with code `OO`? (Hint: `nrow` can be useful here along with logical indexing)

There are `r nrow(nyc[nyc$carrier == "OO",])` flights by the airline with code 00

- How long is the shortest flight out of any NYC airport? (Hint: `min` can be useful, remember to handle `NA` values)

Shortest flight out of NYC airport is `r min(nyc$air_time, na.rm=TRUE)` mins long


- How many flights where there by United Airlines (code: UA) on Jan 12th 2013?

There were `r nrow(nyc[nyc$carrier =="UA" & nyc$month == "1" & nyc$day == 12, ])` flights by UA on Jan 12th 2013

### Arrival Delay


```{r}
janflight = nyc[nyc$month == "1",] #Filtering Jan details

```

- What was the average arrival delay for all airports and all airlines combined in Jan 2013? 

Average arrival delay for all airports and all airlines combined in Jan 2013 is `r round(mean(janflight$arr_delay, na.rm=TRUE), 2)`

- Whats was the median arrival delay for all airports and all airlines combined in Jan 2013?

Median arrival delay for all airports and all airlines combined in Jan 2013 is `r round(median(janflight$arr_delay, na.rm=TRUE), 2)`

- Based on your answers to the two questions above, what can you say about the distribution of arrival delays? 

While most airlines and flights arrive ahead of time (hence median is negative), when flights do get delayed, they get delayed by a huge amount which is pulling the average delay greater than 0


### Airline Performance

- Calculate average arrival delays by airline

```{r}
#Calculating delays by airline
delay = round(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE),2)
```

Average arrival delays by airline as as per below
```{r}
data.frame(delay)
```


- Draw a Bar Plot of Average Arrival Delays for all the Airlines 

```{r}
barplot(delay)
```

- Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay? 

Airline with the highest average delay is `r names(delay[delay==max(delay)])` and the smallest average delay is `r names(delay[delay==min(delay)])`. Airlines with negative average delay are `r names(delay[delay<0])`

### Air Gain

```{r}
#Creating column Air Gain
nyc$airgain <- nyc$dep_delay - nyc$arr_delay
```


a) Explore airgain data - calculate suitable descriptive statistics and appropriate graphics to better understand this data. This part is open ended - you do what you feel works best for you.

```{r}
library(ggplot2)
summary(nyc$airgain)
#Examining airgain for different airlines
airgain_airline <- round(tapply(nyc$airgain, nyc$carrier, mean, na.rm=TRUE),2)
#Examining relationship between distance & airgain - airgain is widely distributed with increase in distance
baseplot <- ggplot(data=nyc, aes(x=distance, y=airgain))
baseplot + geom_point(size=0.7, alpha=0.3)
```

b) 
  - do airlines actually gain any time when in air on average? 
  
  Average airgain is `r round(mean(nyc$airgain, na.rm=TRUE),2)` and `r ifelse(round(mean(nyc$airgain, na.rm=TRUE),2)>0, "Yes, airlines gain time in air on average", "Yes, airlines do not gain time in air on average")`

  - Calculate average airgain for different airlines - which airlines do a better job, which do a worse job?
  
Airlines that gain time on average (most gain to least gain) are `r names(airgain_airline[airgain_airline>0])`
Airlines that lose time on average (most negative to least negative) are `r names(airgain_airline[airgain_airline<0])`

```{r}
#Airlines in decreasing order of airgain
barplot(sort(airgain_airline,decreasing=TRUE))
```


### Merging Data Frames


a) Read the `planes.csv` file using `read.csv` command. Do any data cleaning necessary.

```{r}
#Reading planes data and cleaning
planes <- read.csv("planes.csv")
summary(planes)
str(planes)
planes$year[is.na(planes$year)] <- median(planes$year, na.rm = TRUE)
planes$X <- NULL #Record #
planes$speed <- NULL #Majority of observations are NA
planes$type <- as.factor(planes$type)
planes$manufacturer <- as.factor(planes$manufacturer)
planes$model <- as.factor(planes$model)
planes$engines <- as.factor(planes$engines)
planes$engine <- as.factor(planes$engine)
summary(planes)
```

b) Merge the flights data and the planes data using the `merge` command. You should do the merge on the common column named `tailnum`

```{r}
help("merge")
nyc$year <- NULL
nyc_planes <- merge(nyc, planes, by.x = "tailnum", by.y = "tailnum")
summary(nyc_planes)

```

c) Pose 5 interesting questions from this merged dataset

-Manufacturers used by airlines and their count

```{r}

manufacturer_carrier <- as.data.frame(aggregate(nyc_planes$tailnum ~ nyc_planes$manufacturer + nyc_planes$carrier, data=nyc_planes, length))
manufacturer_carrier
```
-Age of airplanes for each carrier

```{r}
nyc_planes$yearbucket = "<1960"
nyc_planes$yearbucket[nyc_planes$year > 1960 & nyc_planes$year <1970] = "1960 - 1970"
nyc_planes$yearbucket[nyc_planes$year > 1970 & nyc_planes$year <1980] = "1970 - 1980"
nyc_planes$yearbucket[nyc_planes$year > 1980 & nyc_planes$year <1990] = "1980 - 1990"
nyc_planes$yearbucket[nyc_planes$year > 1990 & nyc_planes$year <2000] = "1990 - 2000"
nyc_planes$yearbucket[nyc_planes$year > 2000] = ">2000"
age_flights <- as.data.frame(aggregate(nyc_planes$tailnum ~ nyc_planes$yearbucket + nyc_planes$carrier, data=nyc_planes, length))
age_flights
```
- Average size of flights made by manufacturers as a function of size

```{r}
manufacturersize <- round(tapply(nyc_planes$seats, nyc_planes$manufacturer, mean, na.rm=TRUE),2)
sort(manufacturersize, decreasing = TRUE)
```

- Top 10 manufacturers flying out of NYC

```{r}
manufacturercount <- tapply(nyc_planes$tailnum, nyc_planes$manufacturer, length)
head(sort(manufacturercount, decreasing = TRUE),10)
```

-Year of manufacture of flights from different NYC airports

```{r}
age_origin <- as.data.frame(aggregate(nyc_planes$tailnum ~ nyc_planes$yearbucket + nyc_planes$origin, data=nyc_planes, length))
age_origin

```